Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-20 paper.tex
update intro

[great point from Wong et al] Note that it is important to distinguish 
here between structures and models: whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.
What we want to achieve with this paper is to bring the two concepts closer 
to each other. Compute a likelihood for an approximate (define) ARG under a 
approximate model of the coalescent. 
Approximate in two senses: SMC + leave out precise recombination time info

@@

The breakthroughs in scale achieved by recent methods 
are all based, in different ways, on inferring approximate 
structures instead of a complete and fully detailed history
% in particular with respect to recombination nodes

% first major step forwards in ARG inference
% explicitly outputs an ARG
ARGweaver, generates these ARGs explicitly again using MCMC sampling, by discretizing time 
they can apply HMM-based algorithms to efficiently generate new samples from the distribution of 
possible ARGs.
% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under the SMC, musn't forget to include this.
% model-based approaches vs

% recent advancements: heuristics
Recently, major advancements relying on heuristic rather than model-based approaches
have made ARG inference scalable.
\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
zhang_biobank-scale_2023, zhan_towards_2023}. These methods
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit.
This has also meant however that
we are now using the term ARG for a varying range of approximate structures
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, The degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively. 
[TO DO:] discuss: each of these tools will deal with or resolve uncertainty 
in a different way.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).\\

Recombination in particular:
In contrast to model-based inference methods like \argweaver, 
these models only deal with recombination in an implicit sense.\\

[TO DO:] what does *general* mean? and when in the paper do we discuss it?

[TO DO: finish this section] For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
structures.
Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
under Hudson. \citep{wong_a-general_2023}

% then onto the SMC
[TO ADD: discuss SMC] mention the SMC \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]
Here, the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic information.

Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
and formulate an algorithm 
that is general in the sense that it can deal with the various levels of 
completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
at reconstucting genome-wide genealogies.

@@

Recombination nodes are one such type of unary nodes. Figure \ref{fig:smc-unary} (bottom)
reveals however that all internal nodes but node 4 are unary along some part of 
the genome. These nodes all represent coalescence events.

Note that the presence of these unary nodes allows us to reconstruct the full extent 
of each lineage going back through time. Similarly, all lineages involved in a recombination 
event can be uniquely identified by a change in parent going from one tree to the next 
for the child node associated with that lineage. Both the ARG and the sequence of 
local trees contain the same information, as long as they are simplified to the 
same degree.\\

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows.png}
\caption{ARG (top left) generated under the SMC and its representation as a 
series of local trees. The presence of the unary nodes in the local trees 
ensure these two representations contain the same information. 
The ARG is augmented with haplotype information, coloured according to the 
samples that inherit from it. 
Note how the presence of a unary node on a local tree is 
indicative of either a past (to the left of the local tree) 
or future (to the right) coalescence event. It therefore modifies the 
left-to-right logic of the SMC whereby the floating lineage coalesces randomly
with a the remaining portion of the local tree. 
}
\label{fig:smc-unary}
\end{figure}

@@

% polytomies
[TO DO: polytomies] The above algorithm treats each edge independently which implies 
we can deal with polytomies.
[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
we were able to associate a single likelihood value with each edge in the ARG. 
The compatibility of the algorithm with the \texttt{tskit} library allows us to 
exploit its associated efficiencies both for the current work and its future 
applications. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Thirdly, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of a 
demography-aware genealogy sampler that would allow us to incorporate the information 
provided by recombination events.
More concretely, given our algorithm, any inferred ARG can be used a good starting point 
for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
to the recombination event we further restrict the size of the ARG space that needs to 
be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}). 
$N_e$-changes as well as multiple populations and migration can be readily incorporated 
in the current algorithm.\\

% 2/ hybrid ARG-inference
Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
very recent past and coalescent simulations for the remaining time are a tried and tested 
approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale 
simulations while avoiding the documented biases of the coalescent relative to the 
Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach 
could be used for ARG-inference where one can rely on a heuristic inference method for 
the recent past, while using a model-based approach beyond that cutoff point.\\

% 3/ quantifying uncertainty [unfinished]
Both of these future directions tie in with the desired ability of any ARG-related method to 
quantify the uncertainty associated with

to quantify uncertainty by sampling 
from an underlying distribution.

>>>>

[TL:DR] Structure the introduction around this great point from Wong et al.:
Note that it is important to distinguish here between structures and models: 
whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.
What we want to achieve with this paper is to compute a likelihood for 
an approximate (define) ARG under a approximate model of the coalescent. 
Approximate in two senses: SMC + leave out precise recombination time info.
Goal of the introduction: bring across the need to do this.

@@

%Although the coalescent with recombination best (?) probabilistic description of 
%the events generating an ARG for a given sample, ...

All these: fully precise ARG under the coalescent with recombination.
Computationally prohibitive to explore ARG space:
there are infinitely many ARGs compatible with any genomic dataset, 
and many of these ARGs will have a comparable likelihood \citep{mcvean_approximating_2004}.
By dropping the fully precise requirement that (scalable) ARGs became computationally 
feasible. First ARGwaver.
% ARGweaver, model-based yet approximate (discretised)
% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under 
% the SMC, musn't forget to include this.
Model-based (an approximation of the coalescent, SMC) + approximate in the sense of discretized.
and later really scalable methods:\\

[TL;DR] next step: heuristic methods
Recently, major advancements relying on heuristic rather than model-based approaches
have made ARG inference scalable.
\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
zhang_biobank-scale_2023, zhan_towards_2023}. These methods
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit.
% however, scalability has impacted our ability to connect these to any model
This has also meant however that
we are now using the term ARG for a varying range of approximate structures
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).\\

[TL;DR] Specify  different structural approximations:
1/ Recombination is implicit. Recombinations are not modelled or inferred explicitly, 
they can be observed however as tree changes. However a tree change might be the result of one or 
more recombination events.
2/ each of these tools will deal with or resolve uncertainty 
in a different way.\\

[TL;DR] Specify different model-based approximations:
Coalescent with Recombination as the dominant model to describe ancestry.
Through time different approximations. Wiuf and Hein demonstrating a left-to-right 
algorithm. No longer Markovian. This idea used by McVean and Cardin: SMC. 
And followups (SMC', SMC(k)) \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]\\

[TL;DR] Wrap up by making clear that we will combine the two types of approximations:
Here, the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic information.
Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
and formulate an algorithm 
that is \textit{general} in the sense that it can deal with the various levels of 
completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
at reconstucting genome-wide genealogies.

@@

Recombination nodes are one such type of unary nodes. Figure \ref{fig:smc-unary} (bottom)
reveals however that all internal nodes but node 4 are unary along some part of 
the genome. These nodes all represent coalescence events. Note that the presence of these 
unary nodes allows us to reconstruct the full extent 
of each lineage going back through time as well as uniquely identify all lineages involved 
in a recombination event. Both the ARG and the sequence of 
local trees contain the same information, as long as they are simplified to the 
same degree.\\

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
\caption{ARG (top left) generated under the SMC and its representation as a 
series of local trees. The presence of the unary nodes in the local trees 
ensure these two representations contain the same information. 
The ARG is augmented with haplotype information, coloured according to the 
samples that inherit from it. 
Note how the presence of a unary node on a local tree is 
indicative of either a past (to the left of the local tree) 
or future (to the right) coalescence event. It therefore modifies the 
left-to-right logic of the SMC whereby the floating lineage coalesces randomly
with a the remaining portion of the local tree. 
}
\label{fig:smc-unary}
\end{figure}

@@

% polytomies
[TO DO: polytomies] The above algorithm treats each edge independently which implies 
we can deal with polytomies.
[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]
[TO DO: recombination]: We do not *explicitly* assume a single recombination breakpoint to have 
occurred at each breakpoint.
[TO DO:] On a similar note and more general: we can compute the likelihood for any 
valid tree sequence, however, future work to verify to what extent/in which scenarios 
this is actually helpful.

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
we were able to associate a single likelihood value with each edge in the ARG. 
The compatibility of the algorithm with the \texttt{tskit} library allows us to 
exploit its associated efficiencies both for the current work and its future 
applications. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. 
Secondly, inferred ARGs can be used as a good starting point for a future MCMC sampler.
Thirdly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Finally, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of

$N_e$-changes as well as multiple populations and migration can be readily incorporated 
in the current algorithm.\\

% 2/ great starting point for MCMC
Any inferred ARG can be used a good starting point 
for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
to the recombination event we further restrict the size of the ARG space that needs to 
be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
[TO DO:]

% 3/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
very recent past and coalescent simulations for the remaining time are a tried and tested 
approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale 
simulations while avoiding the documented biases of the coalescent relative to the 
Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach 
could be used for ARG-inference where one can rely on a heuristic inference method for 
the recent past, while using a model-based approach beyond that cutoff point.\\

% 4/ quantifying uncertainty [unfinished]:
Both of these future directions tie in with the desired ability of any ARG-related method to 
quantify the uncertainty associated with
